---
title: "R Bootcamp - List of Functions"
author: "Tu Tran und Andy Gubser"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
  html_document:
    theme: flatly
    toc: true
    toc_depth: 4
    toc_float: 
      collapsed: true
      smooth_scroll: true
    number_sections: true
    df_print: paged

---

\tableofcontents

# R Markdown

```{r globalOpts, include=FALSE, purl=FALSE}
library(knitr)
opts_chunk$set(concordance = TRUE,
               fig.align = "center",
               out.width = "0.7\\linewidth", ## default is 1
               tidy = FALSE,
               comment = NA,
               message = FALSE,
               fig.pos = "H",
               crop = hook_pdfcrop) ## changes mar defaults...
##
options(digits = 2)
```


```{r}
knitr::opts_chunk$set(echo = TRUE, include=TRUE)
```

- The Definitive Guide: https://bookdown.org/yihui/rmarkdown/
- R Markdown Cheat Sheet: https://rstudio.com/wp-content/uploads/2016/03/rmarkdown-cheatsheet-2.0.pdf
- R Markdown Reference Guide: https://rstudio.com/wp-content/uploads/2015/03/rmarkdown-reference.pdf


# Base R
## Global Environment

```{r}
ls() # list all elements in global environment

?mean
help(mean)

apropos("filter")
apropos("filter", where = TRUE)
pos.to.env(4) ## or search()
pos.to.env(6)
```

## Lists
```{r}
seq(from = 1, to = 10, length.out = 5)
rep(x = 1:3, times = 4)
```



## Matrices

```{r}
# create a matrix
M.1 <- matrix(data = 6:1, nrow = 2, ncol = 3)
M.1

# equivalent to
rbind(c(6,4,2), c(5,3,1))
cbind(1:2, 3:4, 5:6)

# add a column to existing matrix
cbind(M.1, c(99, 102))

# replace()
```

## Data Frames
```{r}
## creating a data frame
v.participant.name <- c("Max", "Tony Müller", "Andreas", "Ramona", "Nick Q.")
v.age <- c(17, 23, 45, 11, 44)
v.minor.YES <- c(TRUE, FALSE, FALSE, TRUE, FALSE)
d.test <- data.frame(v.participant.name, v.age, v.minor.YES)
d.test

## class of the single elements
class(v.participant.name)
class(v.age)
class(v.minor.YES)
class(d.test)

## Attributes
colnames(d.test)
rownames(d.test)
# tibble::rownames_to_column(d.sports)


```

## Functions
```{r}
help(mean)
?mean

my.mean <- function(x){
  N <- length(x)
  Sum <- sum(x)
  computed.mean <- Sum / N
  return(computed.mean)
}

v.1 <- rnorm(n = 10^3, mean = 2)
v.2 <- rnorm(n = 10^3, mean = 2)

my.mean(v.1)


```


## Methods

```{r}
lm.iris <- lm(Sepal.Width ~ Species, data = iris)
summary(lm.iris)

methods(summary)
methods(class = "lm")
methods(class = "matrix")
methods(write.csv) # not a generic function


```


# Import Data
```{r}
## set and get working directory
setwd(getwd())

## import data
d.blueEggs <- read.table("../DataSets/BlaueEier.txt",
           header = TRUE)
d.sports <- read.table("http://stat.ethz.ch/Teaching/Datasets/sport.dat",
                       header = TRUE)

## export data
# write.csv2(d.sports, "SportsDataset.csv")
# write.csv()
# write.table()
```

```{r}
## inspect data
str(iris)
dim(iris) ## nrow(iris); ncol(iris)
head(iris)
summary(iris)
table(iris[, "Species"])

min(iris$Sepal.Length)
max(iris$Sepal.Length)
mean(x = 1:13)
range(iris$Sepal.Length)
levels(iris$Species)
nlevels(iris$Species)
length(unique(iris$Sepal.Length))
## see also n_distinct() in package {dplyr}

```


# Data Preprocessing

## Mutate, Filter, Arange, Select with {dplyr}
```{r}
library(dplyr)
d.sports <- tibble::rownames_to_column(d.sports)

d.sports %>% 
  mutate(weit.in.cm = weit*100) %>%
#  select(d.sports, contains("e"))
#                  starts_with(...)
#                  ends_with(...)
  filter(hoch > 200, weit.in.cm >= 780) %>% # filter observations
  arrange(hoch, desc(weit.in.cm)) %>% # sort observations
  select(rowname, weit.in.cm, hoch, punkte) %>%  # select variables
  head

## sample observations
## select random rows from a table
head(d.sports)
sample_n(d.sports, size = 10)
sample_frac(d.sports, size = 0.2)
```

### Replace NA with {tidyr}
```{r}
d.sports.NA <- d.sports
d.sports.NA[1:2, "weit"] <- NA 
d.sports.NA[3:6, "kugel"] <- NA 
d.sports.NA


library(tidyr)
replace_na(d.sports.NA, 
           list(weit = 2, kugel = 10)) %>% 
  head


```

# Export Data
```{r}
saveRDS(d.sports, file = "DSports.RDS")
```


# Data Analysis

## Deskriptive Analysis

```{r}
## summarise data
head(iris)
iris %>% 
  summarise(mean.SepalLength = mean(Sepal.Length))

iris$Sepal.Length %>% ## with pipe
  mean %>% 
  round(digits = 2)

iris %$% ## with exposition pipe
  mean(Sepal.Length) %>% 
  round(digits = 2)

## summarise data with Scaling and Centering of Matrix-like Objects
iris %>% 
  subset(Species == "setosa") %$% 
  scale(Petal.Length, scale = TRUE, center = FALSE) %>% 
  mean


## summarise per group
iris %>% 
  group_by(Species) %>% 
  summarise(mean.SepalLength = mean(Sepal.Length))


## get top n observations
iris %>% 
  group_by(Species) %>% 
  top_n(n = 2, wt = Sepal.Length) ## by default wt is the last column


## counting observations
iris %>% 
  group_by(Species) %>% 
  summarise(No.Obs = n())


## adding a counter
iris %>% 
  mutate(id = 1:n())
## adding a counter within group
iris %>% 
  group_by(Species) %>% 
  mutate(id = 1:n())


## forcing printing
iris %>% 
  group_by(Species) %>% 
  mutate(id = 1:n()) %>% 
  print(width = Inf, n = 51)

## reshape datasets

library(tibble) ## for rownames_to_column()
d.sport <- d.sport %>% 
  rownames_to_column(var = "Athlete")
d.sport

## from wide to long
library(tidyr) ## for gather()
d.sport.long <- gather(d.sport, 
                 key = "discipline", ## new column with name of the discipline
                 value = "result", ## new column with value (unquoted also works)
                 -Athlete) ## variable(s) that is(are) not to put as results
##
head(d.sport.long)
d.sport.long %>% 
  subset(subset = Athlete == "OBRIEN")


## from wide to long
d.sport.wide.again <- spread(d.sport.long, 
                             key = "discipline", ## name from long df
                             value = "result") ## name from long df
head(d.sport.wide.again)

## change from upper case names to lower case names
d.sport.wide.again$Athlete <- tolower(d.sport.wide.again$Athlete)
head(d.sport.wide.again)



```

**Note:** the data format needed for supervised problems is long-format (each row represents a single observation). 
Whereas, for unsupervised problems, wide-format is required (i.e. one represents several measurements). For example, a sample of wine where several aspects where quantified.


## Joining Datasets
```{r}
set.seed(14)
d.age <- data.frame(age = runif(n = 7, min = 19, max = 34),
                    Athlete = d.sport$Athlete,
                    Gender = c("M","M","F","M","F","F","F"))
d.age

library(dplyr)
d.sport.long.age <- left_join(d.sport.long, d.age, by = "Athlete")
head(d.sport.long.age)

class(d.sport.long$Athlete)
class(d.age$Athlete)

```



## Regression Analysis

```{r}
## Generic and methods functions
lm.1 <- lm(Sepal.Length ~ Petal.Length,
           data = iris)
summary(lm.1)
summary(iris)
## same function , different behaviour?
## --> summary is a generic function
class(lm.1)
methods(summary)
methods(class = "lm")

?summary.lm ## has its own help page
?summary.data.frame ## redirects you to the *generic* function help page

```






# Data Visualisation


## Base R plot function
```{r}
d.Puromycin.plot <- plot(y = Puromycin$rate, 
     x = Puromycin$conc,
     pch = 19, # pch 19 is a filled circle
     col = "red", # set color
     cex = 3, # set point size
  
     # add a title and axis labels
     main = "Reaction rates vs. Conc",
     xlab = "substrate conc [ppm]",
     ylab = "reaction rates [counts/min]", 
     type = "p")
# type = "l" # lines
# type = "b" # both (points and lines)
# type = "n" # nothing
# lty = "dashed") # line types

## formula interface
# plot(rate ~ conc, data = Puromycin)

# graphical parameters
graphical.parameters <- par()
head(graphical.parameters)
str(graphical.parameters)

# multiple plots on one screen 
par(mfrow = c(1, 2))
plot(1:19, 1:19, pch = 1:19, col = 1:19)
boxplot(Petal.Length ~ Species, data = iris)

## get back to default values
# dev.off()

## interactive functions ##
## identify
plot(Agriculture ~ Fertility, data = swiss)
identify(y = swiss$Agriculture, x = swiss$Fertility)
## click on "finish" icon
swiss[c(46,19,7), ]

## pannelling ##
head(Loblolly,  n = 8)
levels(Loblolly$Seed)
nlevels(Loblolly$Seed)
table(Loblolly$Seed)

plot(height ~ age, data = Loblolly, col = Seed)

```

## Package {lattice}

```{r}
library(lattice)
xyplot(height ~ age, data = Loblolly,
       type = c("p", "smooth", "g"))
# "p": points
# "b"/"o": both (overplotted)
# "r": add linear regression line
# "smooth": add loess fit
# "spline": add cubic smoothing spline line

```

## Package {ggplot2}

```{r, include=TRUE}

## {ggplot2} ##
library(ggplot2)

## base call ggplot()
ggplot(data = Loblolly,
       mapping = aes(y = height,
                     x = age)) +
  geom_point()

## i) panelling
ggplot(data = Loblolly,
       mapping = aes(y = height,
                     x = age)) +
  geom_point() +
  facet_wrap(. ~ Seed) ## !

# equivaltent to:
qplot(y = height, x = age, data = Loblolly, facets = ~ Seed)


## ii) grouping (colouring)
ggplot(data = Loblolly,
       mapping = aes(y = height,
                     x = age,
                     colour = Seed)) + ## !
  geom_point()
##
ggplot(data = Loblolly,
       mapping = aes(y = height,
                     x = age,
                     colour = Seed)) +
  geom_line() ## !

## ii) grouping (no colouring)
ggplot(data = Loblolly,
       mapping = aes(y = height,
                     x = age,
                     group = Seed)) + ## !
  geom_line()

## iii) adding summary statistics
ggplot(data = Loblolly,
       mapping = aes(y = height,
                     x = age)) +
  geom_point() +
  geom_smooth()
##
ggplot(data = Loblolly,
       mapping = aes(y = height,
                     x = age)) +
  geom_point() +
  geom_smooth(method = "lm")

```

Here the link to the {ggplot2} Cheatsheet: https://github.com/rstudio/cheatsheets/blob/master/data-visualization-2.1.pdf

# Regression Models

```{r}




```



# Packages
Load all packages at the very beginning of your code to avoid surprises and to get the function you want (e.g. base::filter() or mgcv::gam()


```{r, include=FALSE}



## "Default R packages" ##
# sessionInfo()
# find("lm")
# #?lm
# 
# ## add-on packages ##
# ## installing add-on packages
# #install.packages("boot")
# 
# ## loading add-on packages
# library(gam)
# 
# ## help about a package
# #?gam
# #help(package = "gam")
# 
# ## Where are packages installed? ##
# .libPaths()
# 
# ## finding functions? ##
# find("boxplot")
# apropos("packages")
# 
# ## coflicts ##
# ## To make sure that you work with a given function ##
# gam.3 <- mgcv::gam(Sepal.Width ~ s(Petal.Width), data = iris)
# summary(gam.3)
# 
# ## add-on packages evolve
# packageVersion("lme4")

```



# Best practices in Coding

## Chunk structure
1. say what you will do (and why)
2. do it
3. check that it worked
4. comment the results

```{r}
## Example:
## we create a new variable that contains the results of "weit" expressed in cm
d.sports$weit.in.cm <- d.sports$weit
d.sports %>% select(contains("weit")) %>% sample_n(10)
```


## Reproducibility
- rmarkdown/knitr files
- checkpoint package
- git


## Coding conventions

Naming Objects:
- d.xxx for data frame (e.g. d.sports)
- v.xxx for vectors 
- t.xxx for temporary objects (e.g. in for loops)
- lm.xxx for linear models objects
- mod.xxx for other models (or e.g. svm.xxx for SVMs)
- mod.sport.1 is usually less complex than mod.sport.2
- if your current model is e.g. mod.sport.4 and you compute say confidence intervals, name them ci.sport.4

## overwriting
- y <- log(y) # worse bit of code you can write
- y.log <- log(y) # much better
- d.sports.weit.10 or d.sports.short <- d.sports %>% filer(weit >= 10)
- if adding variable to a data frame overwriting is ok
- d.sports <- d.sports %>%mutate(weit.in.cm = weit * 100)

