---
title: "Assignment R Bootcamp"
author: "Tu Tran und Andy Gubser"
date: /today
output: 
  html_document:
    theme: flatly
    toc: true
    toc_depth: 4
    toc_float: 
      collapsed: true
      smooth_scroll: true
    number_sections: true
    df_print: paged

---

```{r setup}
knitr::opts_chunk$set(echo=TRUE)

# install
# install.packages("tinytex")
# tinytex::install_tinytex()


```


# Clarify question

This project analysis the usage of sharing bikes in New York City in 2016 regarding the following properties: 
- Who is using the bikes?
  - gender (m/f) **
  - age (recode from birth year =  2016-x) 
- When are the bikes used
  - time (day/night)
  - date (weekdays/weekend)
  - month (summer/winter) **
  - weather **
  
** Andy tries to include these into the visualisation

- Who is using the bikes for the longest rides?

# Get data
-- # Data prep

# Data visualisation

```{r}
#our background map
# mad_map <- get_map(getbb("New York"),maptype = "toner-background", source="stamen")

min_lon <- min(d.bike$start_station_longitude, d.bike$end_station_longitude)
max_lon <- max(d.bike$start_station_longitude, d.bike$end_station_longitude)
min_lat <- min(d.bike$start_station_latitude, d.bike$end_station_latitude)
max_lat <- max(d.bike$start_station_latitude, d.bike$end_station_latitude)
mad_map <- (map <- get_map(c(left = min_lon, bottom = min_lat, right = max_lon, top = max_lat)))

#final map
ggmap(mad_map)+ 
  
    geom_leg(data=d.bike.sample,
      color = gender, 
      alpha=0.1,
      
      aes(
        x = start_station_longitude, 
        y = start_station_latitude, 
        xend = end_station_longitude, 
        yend = end_station_latitude
        )
      )+
  
  facet_grid(month ~ gender) +
  theme(
    axis.title.x=element_blank(),
    axis.text.x=element_blank(),
    axis.ticks.x=element_blank(),
        
    axis.title.y=element_blank(),
    axis.text.y=element_blank(),
    axis.ticks.y=element_blank()
    ) + 

  labs(x="",y="")

# ggsave("bike_nyc", device = "pdf")

# y-var: tripduration
# x-var: wheather, gender, age, month, time



```














```{r}

ggmap(ggmap, source="osm") + 
  geom_leg(
    aes(x = start_station_longitude, 
        y = start_station_latitude, 
        xend = end_station_longitude, 
        yend = end_station_latitude,
        colour = "blue", alpha=0.0001
    )
  )


nyc_map <- fortify(gSimplify(nyc, 0.05))

gg <- ggplot()
gg <- gg + geom_map(data=nyc_map, map=nyc_map,
                    aes(x=long, y=lat, map_id=id),
                    color="black", fill="white", size=0.25)
gg <- gg + coord_equal() 
gg <- gg + theme_map()
gg



```


```{r}
colnames(d.bike)


```

















```{r}
# Set value for the minZoom and maxZoom settings.
leaflet(options = leafletOptions(minZoom = 0, maxZoom = 18))

m <- leaflet() %>%
  addTiles() %>%  # Add default OpenStreetMap map tiles
  addMarkers(lng=d.bike$start_station_longitude, 
             lat=d.bike$start_station_latitude, 
             popup="The birthplace of R")
m  # Print the map

```

d.bike



# Fit models




# Interpretation





```{r}
# sessionInfo()
# {miniCRAN} pkgDep()

```


# https://rstudio.github.io/leaflet/