---
title: "Assignment R Bootcamp"
author: "Tu Tran and Andy Gubser"
date: /today
output: 
  html_document:
    theme: flatly
    toc: true
    toc_depth: 4
    toc_float: 
      collapsed: true
      smooth_scroll: true
    number_sections: true
    df_print: paged

---

# Clarify question

# get data: d1 <- read_csv(relative path)
# data prep: saveRDS(d1, "file.RDS")
# data visualisation: set.seed(19)
# fit models
# interpretation


# Abstract

# Purpose / Research question

In this project we analyse the typical subscriber of bike sharing services in New York City. In particular, we evaluate who made the longest trips. Therefore, we identify subscibers by their age and gender as well as the season, the weekday and the daytime of the ride. We further consider the weather conditions during the ride. 

# Methodology
The analysis bases on two data sets from Kaggle: the bikeshare data set for New York City in 2016 [^1 https://www.kaggle.com/samratp/bikeshare-analysis#NYC-CitiBike-2016.csv] and the corresponding weather data set [^2 https://www.kaggle.com/mathijs/weather-data-in-new-york-city-2016]. 

After data preparation, we provide an graphical analysis to approach the research question of who is doing the longest trips. Then, we fit a linear model followed by a random forest model to deepen our analysis.  

# Analysis

## R and knitr Setup
```{r setup}
# remove all objects loaded and clear memory
rm(list = ls(all.names = TRUE))
gc()
library(checkpoint)
checkpoint(snapshotDate = "2020-12-29")

knitr::opts_knit$set(root.dir = rprojroot::find_rstudio_root_file())
knitr::opts_chunk$set(echo=TRUE)
set.seed(19)

```


## Library Import

```{r importLibraries, include=FALSE}
list.of.packages <- c("installr", "Hmisc", "ggmap", "tidyverse", "viridis", "leaflet", "lubridate")
new.packages <- list.of.packages[!(list.of.packages %in% installed.packages()[,"Package"])]
if(length(new.packages)) install.packages(new.packages)
for (p in list.of.packages){
  library(p, character.only = TRUE)
}

```

# Data Import

```{r dataImport, include=FALSE, echo=FALSE, }
# # preprocess bike data
# source("./d.bike_preprocessing.R")

# # preprocess weather data and merge with preprocessed bike data
# source("./d.weather_preprocessing.R")

d.bike_weather <- readRDS("../data/d.bike_weather.rds")

colnames(d.bike_weather)
summary(d.bike_weather)
dim(d.bike_weather)
len(d.bike_weather)

```


## Data Definition

The bike data set consists of 276'798 observation and 15 variables. These variables include the two categories user type and gender as well as the numeric birth year. User type defines if the user is a registered subscriber of the bike sharing service or a casual user, whereby we focus on subscribers. Gender says if the user is female, male or unknown. Since there are only a few user with unknown gender, we exclude them from the analysis. The continous variable age is calculated from the birth year. Further, the categorical variable age_cor is created, where users over 70 are put into a single bin in order to reduce potential bias of these observations. 
Further, the data set include start and stop time stamps as well as trip duration of individual trips, station names, ids and its coordinates. By checking for outliers, one observation was detected to have meaningless coordinates. These observations are excluded from further analysis. 
The prepared data set includes 242'746 observations and 21 variables. 

The weather data set includes 366 observations and the 7 variables maximum, minimum and average temperature, precipitation, snow fall and snow depth on specific dates. In order to merge the two data sets on the date variable, this variable is created in the bike sharing data set by extracting the date from the start timestamps. The merged data set consists of 242'746 observations and 15 variables. 


# Graphical Analysis


```{r}
ggplot(data = d.bike_weather, aes(y = tripduration, x = age, color=gender))+ 
    geom_point()
#   facet_grid(month ~ gender) +

```

























```{r, include=FALSE, echo=FALSE}
# get background map
min_lon <- min(d.bike_weather$start_station_longitude, d.bike_weather$end_station_longitude)
max_lon <- max(d.bike_weather$start_station_longitude, d.bike_weather$end_station_longitude)
min_lat <- min(d.bike_weather$start_station_latitude, d.bike_weather$end_station_latitude)
max_lat <- max(d.bike_weather$start_station_latitude, d.bike_weather$end_station_latitude)

mad_map <- (map <- get_map(c(left = min_lon, bottom = min_lat, 
                             right = max_lon, top = max_lat)))

# take a random sample of size 1%
d.bike_weather_sample = d.bike_weather[
  base::sample(x=1:nrow(d.bike_weather), round(nrow(d.bike_weather)*1/100)), ]


```



```{r}
#final map
#ggmap(mad_map) + 
#  geom_leg(
#    data=d.bike_weather_sample,
#      # color = factor(gender), 
#      alpha=0.1,
#      aes(
#        color = age_corr,
#        x = start_station_longitude, 
#        y = start_station_latitude, 
#        xend = end_station_longitude, 
#        yend = end_station_latitude
#        )
#      )+
#  labs(x="",y="") +
#  facet_grid(month ~ gender) +
#  theme(
#    axis.title.x=element_blank(),
#    axis.text.x=element_blank(),
#    axis.ticks.x=element_blank(),
        
#    axis.title.y=element_blank(),
#    axis.text.y=element_blank(),
#    axis.ticks.y=element_blank()
#    ) + 
#  scale_color_viridis(option = "magma")

#ggsave("map_age_gender_month.pdf", device = "pdf")
```
















```{r}
# Set value for the minZoom and maxZoom settings.
#leaflet(options = leafletOptions(minZoom = 0, maxZoom = 18))

#m <- leaflet() %>%
#  addTiles() %>%  # Add default OpenStreetMap map tiles
#  addMarkers(lng=d.bike_weather$start_station_longitude, 
#             lat=d.bike_weather$start_station_latitude, 
#             popup="The birthplace of R")
#m  # Print the map

```


```{r}




```


# Fit models





# Check Results (output and graphs)





# Interpretation





```{r}
# sessionInfo()
# {miniCRAN} pkgDep()

```
